---
title: "CornFuturePrice-Model"
author: "Seema Rani Kanuri"
date: "September 13, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r}
Monthly_Average_Ratio <-read.csv('D:\\AgReliant\\CropData\\Features\\Monthly_Average_Ratio.csv')
str(Monthly_Average_Ratio)
Monthly_Average_Ratio <- subset(Monthly_Average_Ratio, , -c(X))
```

```{r}
FutureStMillionbushels <-read.csv('D:\\AgReliant\\CropData\\InputData\\Feed_Grains_Excel_stocks.csv')
#iconv(names(FutureStMillionbushels), to='ASCII', sub='')
#FutureStMillionbushels$year <- FutureStMillionbushels$ï..year
names(FutureStMillionbushels) <- gsub("..", "", names(FutureStMillionbushels), fixed = TRUE)
names(FutureStMillionbushels) <- gsub("ï", "", names(FutureStMillionbushels), fixed = TRUE)

```


```{r}
head(FutureStMillionbushels)
```

```{r}
StockFuture <- subset(FutureStMillionbushels, select=c("year","data_type","value"))

```

# Add a new feature: fraction of annual corn supply that is in stockpiles
# at the beginning of each year (possibly associated with decrease in
# corn futures price?)
```{r}


library("reshape2") # for dcast()
StockFuture <- dcast(StockFuture, year ~ data_type,  value.var = "value")
```


```{r}
colnames(StockFuture) <- c('year', 'beginning_stocks', 'total_domestic_use', 'total_supply')
fraction_stockpiled <- as.numeric(StockFuture$beginning_stocks) / as.numeric(StockFuture$total_supply)
avg_fraction_stockpiled <- mean(fraction_stockpiled)
StockFuture[, 'fraction_stockpiled'] <- fraction_stockpiled
StockFuture <- subset(StockFuture, select = +c(year, fraction_stockpiled))
```

```{r}
Merge_output_df6 <- merge(x = Monthly_Average_Ratio, y = StockFuture, by = c('year'))
#write.csv(Merge_output_df6,'D:\\AgReliant\\CropData\\Merge_output_df6.csv',row.names=FALSE)

```

# Modeling to predict the Corn Future price 

```{r}
library(caret)
input11 = read.csv('D:\\AgReliant\\CropData\\Features\\Merge_output_df5.csv')
input12 = Merge_output_df6

df <- merge(x = input12, y = input11, by = c('year'))

vc = seq(1975,2016, by=1)
vc_test = seq(2017,2017, by=1)

input_df_c <- df[df$year %in% vc,]
test_c <- df[df$year %in% vc_test,]

```


```{r}

split_c<-createDataPartition(y = input_df_c$price_per_bushel_december, p = 0.8, list = FALSE)

dev_c<-input_df_c[split_c,]
val_c<-input_df_c[-split_c,]

```

#Model Building and Tuning:
```{r}

lmFit_c <- train(price_per_bushel_december ~ ., data = dev_c, method = "rf")
summary(lmFit_c)

```

#Another useful function would be "trainControl" which allows for estimation of parameter coefficients through resampling methods
#like cross validation, boosting etc.
```{r}

ctrl_c <-trainControl(method = "cv" ,number = 10)
lmCVFit_c <-train(price_per_bushel_december ~ ., data = input_df_c, method = "rf", trControl = ctrl_c, metric= "Rsquared")
summary(lmCVFit_c)

```


#Model Performance

```{r}

residuals_c<-resid(lmFit_c)
predictedValues_c<-predict(lmFit_c)
plot(dev_c$price_per_bushel_december,residuals_c)
abline(0,0)
plot(dev_c$price_per_bushel_december,predictedValues_c)
```

#function "defaultSummary" can be used which in this example returns the values of R-squared and RMSE metrics.
```{r}
predictedval_c<-predict(lmFit_c,val_c)
modelvalues_c<-data.frame(obs = val_c$price_per_bushel_december, pred=predictedval_c)
defaultSummary(modelvalues_c)
```

```{r}

predictedval_c_cv<-predict(lmCVFit_c,val_c)
modelvalues_c_cv<-data.frame(obs = val_c$price_per_bushel_december, pred=predictedval_c_cv)
defaultSummary(modelvalues_c_cv)
```


# Model 2

```{r}

#install.packages("rpart")
library(rpart)
rpart.model <- rpart(price_per_bushel_december ~ .,data = input_df_c, method = "anova")
rpart.model
cost_dec <- predict(rpart.model, test_c)
modelvalues_c17<-data.frame(value=c("price_per_bushel_december"), year = test_c$year, pred=cost_dec*100)
#modelvalues_c17 <- rbind(modelvalues_c17, data.frame(value=c("price_per_bushel_december in Millions"), year = test_c$year, pred=cost_dec*1000000))

```


#simimarly we can check for price_per_bushel_november, price_per_bushel_october ,price_per_bushel_september
```{r}
rpart.model1 <- rpart(price_per_bushel_november ~ .,data = input_df_c, method = "anova")
rpart.model1
cost_nov <- predict(rpart.model1, test_c)
modelvalues_c17<- rbind(modelvalues_c17, data.frame(value=c("price_per_bushel_november"), year = test_c$year, pred=cost_nov*100))

```

```{r}

rpart.model2 <- rpart(price_per_bushel_october ~ .,data = input_df_c, method = "anova")
rpart.model2
cost_oct <- predict(rpart.model2, test_c)
modelvalues_c17<- rbind(modelvalues_c17, data.frame(value=c("price_per_bushel_october"), year = test_c$year, pred=cost_oct*100))


```


```{r}

rpart.model3 <- rpart(price_per_bushel_september ~ .,data = input_df_c, method = "anova")
rpart.model3

```

```{r}
cost_sep <- predict(rpart.model3, test_c)
modelvalues_c17<- rbind(modelvalues_c17, data.frame(value=c("price_per_bushel_september"), year = test_c$year, pred=cost_sep*100))

modelvalues_c17
```

#The average price of corn per bushel in  September, October , November , December The prices shown are in units of USD so 3.622 means 362.2 cents per Bushel.All the future Price tables outstanding contracts open/closed interest are shown in cents per Bu



#References

##Data

###All the data used below is been downaloaded from open source like ;
### Monthly temperature, precipitation ; ftp://ftp.ncdc.noaa.gov/pub/data/cirs/climdiv/ ;   https://data.ers.usda.gov/FEED-GRAINS-custom-query.aspx
###Planting progress ;  https://quickstats.nass.usda.gov/results/0D72DC4F-11B7-35DE-A5B1-D6AA348291F4
###continuous corn futures index ;   https://www.quandl.com/data/CHRIS/CME_C1-Corn-Futures-Continuous-Contract-1-C1-Front-Month
### BBeginning-of-year corn grain stockpiles ; https://data.ers.usda.gov/FEED-GRAINS-custom-query.aspx#ResultsPanel
###Palmer drought severity index definition ; ftp://ftp.ncdc.noaa.gov/pub/data/cirs/climdiv/climdiv-pdsist-v1.0.0-20170906


##Model Implementation
###https://www.analyticsvidhya.com/blog/2014/12/caret-package-stop-solution-building-predictive-models/
###https://www.analyticsvidhya.com/blog/2016/03/practical-guide-principal-component-analysis-python/
###https://www.analyticsvidhya.com/blog/2016/02/complete-tutorial-learn-data-science-scratch/
###http://mindymallory.github.io/PriceAnalysis/commodity-price-analysis-and-forecasting.html 